Query Optimization for Dynamic Imputation

### 2017-05-11 paper/conclusion.tex

\section{Conclusion}
Our evaluation of \ProjectName{} shows that missing values and their imputation can be successfully integrated into the relational calculus and
existing cost-based plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
learning technique, as operators in the algebra and use a simple yet effective cost model to consider trade-offs in 
imputation quality and runtime. We show that different values for the trade-off parameter can yield substantially
different plans, allowing the user to express their preferences for performance on a per-query basis.
Our experiments with the CDC NHANES, freeCodeCamp and ACS datasets
show that \ProjectName{} can be used successfully with real-world data to run standard SQL queries over data with missing values. 
We craft a series of realistic queries, meant to resemble the kind of questions an analyst might ask during
the data exploration phase. The plans selected for each query execute an order-of-magnitude faster than
 the standard approach of imputing on the entire base tables and
then formulating queries. Furthermore, the difference in query results between the two approaches is
shown to be small in all queries considered (0-\highsmapealphaoneexacs\% error, \Cref{sec:results}). Data analysts need not commit to a specific imputation strategy and can instead
vary this across queries.

>>>>

\section{Conclusion}
Our evaluation of \ProjectName{} shows that missing values and their imputation can be successfully integrated into the relational algebra and
existing cost-based plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
learning technique, as operators in the algebra and use a simple yet effective cost model to consider trade-offs in 
imputation quality and runtime. We show that different values for the trade-off parameter can yield substantially
different plans, allowing the user to express their preferences for performance on a per-query basis.
Our experiments with the CDC NHANES, freeCodeCamp and ACS datasets
show that \ProjectName{} can be used successfully with real-world data to run standard SQL queries over data with missing values. 
We craft a series of realistic queries, meant to resemble the kind of questions an analyst might ask during
the data exploration phase. The plans selected for each query execute an order-of-magnitude faster than
 the standard approach of imputing on the entire base tables and
then formulating queries. Furthermore, the difference in query results between the two approaches is
shown to be small in all queries considered (0-\highsmapealphaoneexacs\% error, \Cref{sec:results}). Data analysts need not commit to a specific imputation strategy and can instead
vary this across queries.

### 2017-05-11 paper/implementation.tex

\section{Implementation}
Adding imputation and our optimization algorithm to a standard SQL database requires modifications to several key database components.
In this section we discuss the changes which must be made to implement dynamic imputation.

\begin{itemize}
\item \textbf{Extended histograms:}
  Ranking query plans with imputation requires estimates of the distribution of missing values in the base tables and in the outputs of sub-plans.
  We extended standard histograms to include the count of missing values for each attribute.
  On the base tables, this provides an exact count of the missing values.
  In sub-plans we use simple heuristics---as discussed in Section~\ref{sec:cardinal}---to estimate the number of missing values after applying both standard and imputation operators.
  These cardinality estimates are critical to the optimizer's ability to compare the performance and imputation quality of different plans.

>>>>

\section{Implementation}
Adding imputation and our optimization algorithm to a standard SQL database requires modifications to several key database components.
In this section we discuss the changes required to implement dynamic imputation.

\begin{itemize}
\item \textbf{Extended histograms:}
  Ranking query plans with imputation needs estimates of the distribution of missing values in the base tables and in the outputs of sub-plans.
  We extended standard histograms to include the count of missing values for each attribute.
  On the base tables, this provides an exact count of the missing values.
  In sub-plans we use simple heuristics---as discussed in Section~\ref{sec:cardinal}---to estimate the number of missing values after applying both standard and imputation operators.
  These cardinality estimates are critical to the optimizer's ability to compare the performance and imputation quality of different plans.
